---
title: "An introduction to rScudo"
date: "`r Sys.Date()`"
author:
-   name: Matteo Ciciani
    affiliation: &aff CIBIO, Univ. of Trento, Trento, Italy
    email: matteo.ciciani@gmail.com
-   name: Thomas Cantore
    affiliation: *aff
output:
    BiocStyle::pdf_document:
        #toc_float: true
vignette: >
    %\VignetteIndexEntry{Signature-based Clustering for Diagnostic Purposes}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

This package implements in R the SCUDO pipeline proposed in Lauria (2013) and
Lauria, Moyseos, Priami (2015).

SCUDO is a method for the analysis of gene expression profiles for diagnostic
and classification purposes. The method is based on the idea of a
sample-specific gene signatures. The expected result is the emergence of a
partitioning of the set of samples in separate clusters (propely, communities)
on the basis of signature similarity.

First, normalization and feature selection are optionally performed. Feature
selection is carried out through a Wilcoxon rank sum test or through a
Kruskal-Wallis rank sum test depending on the number of groups considered.

The subsequent operations include single sample gene ranking and the extraction
of signatures form up-regulated and down-regulated genes. The sizes of the
signatures are customizable. An all-to-all distance matrix is then computed
using a metric similar to the Gene Set Enrichment Analysis (GSEA): the distance
between two samples is computed as the mean of the enrichment scores (ES) of the
signature of each sample in the expression profile of the other sample.
Consensus signtures are then computed, both for up- and down-regulated genes and
for each group. This provides a list of interesting genes of each group.

Finally, a user-defined threshold N is used to generate a network of samples.
The distance matrix is treated as an adjacency matrix, but only the distances
that fall below the N^th^ percentile of distances are used to draw arcs in the
network. The network can be displayed in R or using Cytoscape.

# Example workflow of rScudo

## Data preparation

In this example we will use the `r Biocpkg("ALL")` dataset, containing gene
expression data from T- and B-cells acute lymphocytic leukemia patients. In this
first part, we are interested in distinguishing B-cells and T-cells samples,
based on gene expression profiles. We begin by loading relevant libraries and
subsetting the dataset, dividing it in a training and a testing set.

```{r, message=FALSE}
library(rScudo)
library(igraph)
library(ALL)
data(ALL)

bt <- as.factor(stringr::str_extract(pData(ALL)$BT, "^."))
summary(bt)

set.seed(1)
inTrain <- c(sample(1:95, 47), sample(96:128, 16))
trainData <- ALL[, inTrain]
testData <- ALL[, -inTrain]
```

## Analysis of the training set

We start by analyzing the training set. We first run `scudoTrain`, which returns
an object of class `ScudoResults`.

```{r}
trainRes <- scudoTrain(trainData, groups = bt[inTrain], nTop = 100,
    nBottom = 100, alpha = 0.1)
trainRes
```

From this object we can extract the signatures for each sample and the consensus
signatures for each group.

```{r}
upSignatures(trainRes)[1:5,1:5]
consensusUpSignatures(trainRes)[1:5, ]
```

The object `trainRes` can be used to generate a network of samples, using the
function `ScudoNetwork`. This function returns an `r CRANpkg("igraph")` object.
The parameter `N` controls the percentage of edges to keep in the network. We
can plot this network using the function `scudoPlot`.

```{r}
trainNet <- scudoNetwork(trainRes, N = 0.2)
scudoPlot(trainNet, vertex.label = NA)
```

You can also render the network in Cytoscape, using the function
`ScudoCytoscape`. Note that Cytoscape has to be open when running the function.

```{r, eval=FALSE}
scudoCytoscape(netTrain)
```

Since we obtained a very good separation of the two groups, we proceed to
analyze the testing set.

## Analysis of the testing set

We can use a `ScudoResults` object and the function `scudoTest` to analyze the
testing set. The feature selection is not performed in the testing set. Instead,
only the features selected in the training are used in the analysis of the
testing set.

```{r}
testRes <- scudoTest(trainRes, testData, bt[-inTrain], nTop = 100,
    nBottom = 100)
testRes
```

We can generate a network of samples and plot it.

```{r}
testNet <- scudoNetwork(testRes, N = 0.2)
scudoPlot(testNet, vertex.label = NA)
```

We can use a community clustering algorithm to identify clusters of samples. In
the following example we use the function `cluster_spinglass` to perform a
greedy clustering of our network. In Cytoscape we can perform a similar analysis
using the clustering functions from the clusterMaker app.

```{r}
testClust <- cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)
```

### Supervised classification

`scudoClassify` performs supervised classification of sample in a testing set
using a model built from samples in a training set. It uses a method based on
neighbors in the graph to assign a class label to each sample in the testing
set. We suggest to use the same `N`, `nTop`, `nBottom` and `alpha` that were
used in the trainig step.

```{r}
classRes <- scudoClassify(trainData, testData, N = 0.2, nTop = 100,
    nBottom = 100, trainGroups = bt[inTrain], alpha = 0.1)
classRes$predicted
```

Classification performances can be explored using the `confusionMatrix` function
from `r CRANpkg("caret")`.

```{r}
caret::confusionMatrix(classRes$predicted, bt[-inTrain])
```

## Example of multigroup analysis

```{r}
bcell <- grep("^B", as.character(ALL$BT))
moltyp <- which(as.character(ALL$BT) %in%
c("B1", "B2", "B3"))
bcrneg <- ALL[, moltyp]
bcrneg$mol.biol <- factor(bcrneg$mol.biol)

res <- scudoTrain(bcrneg, factor(bcrneg$BT), 50, 50, 0.01)
```

```{r}
net <- scudoNetwork(res, 0.15)
scudoPlot(net, vertex.label = NA)
```

# Methods in brief

# Session info

```{r}
sessionInfo()
```

# References

Mario Lauria. Rank-based transcriptional signatures. Systems Biomedicine. 2013;
1(4):228-239.

Mario Lauria, Petros Moyseos and Corrado Priami. SCUDO: a tool for
signature-based clustering of expression profiles. Nucleic Acids Research. 2015;
43(W1):W188-92.
